

\textit{Note: The first part of this lecture is a recall on the Maximum-Subarray Problem. As requested, this summary focuses on the new concepts: Matrix Multiplication and Heaps.}

\subsection{Matrix Multiplication}
\textbf{Problem Definition:} Given two $n \times n$ (square) matrices $A$ and $B$, compute their product $C = A \cdot B$.

\begin{itemize}
    \item \textbf{Naive Algorithm:} Computes the product using three nested loops, calculating $c_{ij} = \sum_{k=1}^{n} a_{ik} b_{kj}$. Running time is $\Theta(n^3)$ and space is $\Theta(n^2)$.
    \item \textbf{Simple Divide-and-Conquer:} Divides matrices into $n/2 \times n/2$ submatrices. It performs $8$ recursive multiplications. The recurrence is $T(n) = 8T(n/2) + \Theta(n^2)$, which still yields $T(n) = \Theta(n^3)$.
    \item \textbf{Strassen's Algorithm:} A smarter divide-and-conquer approach that reduces the number of recursive multiplications from 8 to 7, at the cost of some additional matrix additions.
    \begin{itemize}
        \item Recurrence: $T(n) = 7T(n/2) + \Theta(n^2)$.
        \item Running Time: Using the Master Theorem, $T(n) = \Theta(n^{\log_2 7}) \approx \Theta(n^{2.8})$, which asymptotically beats the naive $\Theta(n^3)$ approach.
    \end{itemize}
\end{itemize}

\subsection{Heaps and Heapsort}
Heapsort is an in-place sorting algorithm that runs in $\mathcal{O}(n \log n)$ worst-case time, combining the best properties of Insertion Sort (in-place) and Merge Sort ($\mathcal{O}(n \log n)$).

\subsubsection{The Heap Data Structure}
A binary heap is a nearly complete binary tree.
\begin{itemize}
    \item \textbf{Array representation:} It can be stored in an array without pointers. For a node at index $i$:
    \begin{itemize}
        \item \textsc{Left}$(i) = 2i$ 
        \item \textsc{Right}$(i) = 2i + 1$
        \item \textsc{Parent}$(i) = \lfloor i/2 \rfloor$ 
    \end{itemize}
    \item \textbf{Max-Heap Property:} For every node $i$ other than the root, the key of its parent is greater than or equal to its own key ($A[\textsc{Parent}(i)] \ge A[i]$). The maximum element is always at the root.
    \item \textbf{Height:} The height of a heap of size $n$ is $\Theta(\log n)$.
\end{itemize}

\subsubsection{Heap Operations}
\begin{itemize}
    \item \textsc{Max-Heapify(A, i)}: Ensures the max-heap property for the subtree rooted at $i$, assuming its left and right subtrees are already max-heaps. It propagates the element down the tree and takes $\Theta(\text{height of } i) = \mathcal{O}(\log n)$ time.
    \item \textsc{Build-Max-Heap(A)}: Constructs a max-heap from an unordered array of length $n$. By a tight analysis (since time is linear in the height of the node), it runs in $\mathcal{O}(n)$ time rather than $\mathcal{O}(n \log n)$.
\end{itemize}

\subsubsection{The Heapsort Algorithm}
\begin{enumerate}
    \item Build a max-heap from the input array.
    \item Starting with the root (the maximum element), place it into its correct position by swapping it with the last element in the array.
    \item "Discard" this last node by decreasing the heap size, and call \textsc{Max-Heapify} on the new root to restore the max-heap property.
    \item \textbf{Total Running Time:} $\mathcal{O}(n \log n)$ because building the heap takes $\mathcal{O}(n)$ and we call \textsc{Max-Heapify} ($\mathcal{O}(\log n)$) $n-1$ times.
\end{enumerate}

\subsection{Priority Queues}
A priority queue maintains a dynamic set of elements where each element has an associated "key". Heaps efficiently implement priority queues. A max-priority queue supports the following operations:
\begin{itemize}
    \item \textsc{Maximum(S)}: Returns the element with the largest key in $\mathcal{O}(1)$ time.
    \item \textsc{Extract-Max(S)}: Removes and returns the maximum element in $\mathcal{O}(\log n)$ time.
    \item \textsc{Increase-Key(S, x, k)}: Increases the key of element $x$ to $k$ and traverses upward to restore the heap property in $\mathcal{O}(\log n)$ time.
    \item \textsc{Insert(S, x)}: Inserts element $x$ into the set in $\mathcal{O}(\log n)$ time.
\end{itemize}