\subsection{The Sorting Problem}

\textbf{Input:} a sequence of $n$ numbers $\langle a_1, a_2, \dots, a_n \rangle$.\\
\textbf{Output:} a permutation $\langle a'_1, a'_2, \dots, a'_n \rangle$ such that
$a'_1 \le a'_2 \le \dots \le a'_n$.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Insertion Sort}

The idea is similar to sorting a hand of cards: we go through the elements one by one and
insert each new element into its correct position among those already sorted.

\begin{center}
\begin{tikzpicture}[font=\small]
  % Array cells
  \foreach \val/\idx in {5/1,2/2,4/3,6/4,1/5,3/6}{
    \node[draw, minimum width=0.9cm, minimum height=0.7cm,
          fill=blue!8] at (\idx*1.0-0.5, 0) {\val};
    \node[font=\tiny, gray] at (\idx*1.0-0.5, -0.55) {\idx};
  }
  % Arrow showing "key"
  \draw[->, keyred, thick] (2.5, -0.9) -- (2.5, -0.4)
        node[midway, right=2pt, keyred] {\textit{key}};
  % Already sorted bracket
  \draw[decorate, decoration={brace,amplitude=5pt}, keygreen]
        (0.05, 0.38) -- (1.45, 0.38)
        node[midway, above=5pt, keygreen] {sorted};
\end{tikzpicture}
\end{center}

\begin{algorithm}[H]
\caption{Insertion-Sort($A$)}
\begin{algorithmic}[1]
\For{$j = 2$ \textbf{to} $A.length$}
    \State $key \leftarrow A[j]$  \AlgComment{element to be inserted}
    \State $i \leftarrow j - 1$
    \While{$i > 0$ \textbf{and} $A[i] > key$}
        \State $A[i+1] \leftarrow A[i]$  \AlgComment{shift to the right}
        \State $i \leftarrow i - 1$
    \EndWhile
    \State $A[i+1] \leftarrow key$  \AlgComment{insert in the correct place}
\EndFor
\end{algorithmic}
\end{algorithm}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Proof of Correctness: Loop Invariant}

To prove that an iterative algorithm is correct, we use a \textbf{loop invariant},
a property verified at three key moments:

\begin{enumerate}
  \item \textbf{Initialization}: true before the first iteration.
  \item \textbf{Maintenance}: if true before an iteration, it remains true before the next one.
  \item \textbf{Termination}: at the end of the loop, the invariant proves correctness.
\end{enumerate}

\textbf{Invariant for Insertion Sort:} at the start of each iteration $j$, the
subarray $A[1 \ldots j-1]$ contains the elements originally in $A[1 \ldots j-1]$,
\emph{but sorted}.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Complexity Analysis (RAM model)}

We use the \textbf{Random-Access Machine (RAM)} model: each simple instruction
(arithmetic, memory access, test) takes a constant time.

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lll}
\toprule
\textbf{Case} & \textbf{Condition} & \textbf{Complexity} \\
\midrule
Best case    & Array already sorted (while never executes) & $\Theta(n)$ \\
Worst case   & Array sorted in reverse order              & $\Theta(n^2)$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Worst-case calculation:}
\[
  T(n) = \sum_{j=2}^{n} (j-1) = \frac{n(n-1)}{2} = \Theta(n^2)
\]

\begin{tcolorbox}[colback=lightblue, colframe=keyblue, fonttitle=\bfseries,
                  title={Why focus on the worst case?}]
It provides an \textbf{upper bound guarantee} on the execution time, valid for any
possible input. This is the standard reference in algorithm analysis.
\end{tcolorbox}